Conservation and losses of avian non-coding RNA loci

### 2014-05-29 paper/manuscript.tex
Added Tammy\47s edits.

\begin{abstract}
Here we present the results of a large-scale bioinformatic annotation
of non-coding RNA loci in 48 avian genomes. Our approach uses
probabilistic models of hand-curated families from the Rfam database
to infer conserved RNA families within each avian genome. We
supplement these annotations with predictions from the tRNA annotation
tool, tRNAscan-SE and microRNAs from miRBase.  We show a number of
lncRNA-associated loci are conserved between birds and mammals,
including several intriguing cases where the reported mammalian lncRNA
function is not conserved in birds.  We also demonstrate extensive
conservation of classical ncRNAs (e.g., tRNAs) and more recently
discovered ncRNAs (e.g., snoRNAs and miRNAs) in birds. Furthermore, we
describe numerous ``losses'' of several RNA families, and attribute
these to genuine loss, divergence or missing data.  In particular, we
show that many of these losses are due to issues with assembling
microchromosomes using next-generation sequencing methods. These
combined results illustrate the utility of applying homology-based
methods for annotating novel vertebrate genomes.
\end{abstract}

@@

New sequencing technologies have dramatically expanded the rate at
which ncRNAs are discovered and their functions are determined
\cite{cech2014noncoding}. However, in order to determine the full
range of ncRNAs across multiple species we would require multiple RNA
fractions (e.g. long and short), in multiple species, in multiple
developmental stages and tissues types. The costs of this approach are
still prohibitive in terms of researcher-time and
finances. Consequently, for this work we have concentrated on
bioinformatic approaches. However, we do validate the majority of
these predictions using RNA-seq. Primarily, homology based methods,
namely covariance models (CMs). These remain state of the art for
ncRNA bioinformatic analyses
\cite{Sakakibara:1994,Eddy:1994,Nawrocki:2009} and have well
established sensitivity and specificity rates \cite{Freyhult:2007}.
The CM based approach for annotating ncRNAs in genomes requires
reliable alignments and consensus secondary structures of
representative sequences of RNA families, many of which can be found
at Rfam
\cite{Griffiths-Jones:2003,Griffiths-Jones:2005,Gardner:2009,Gardner:2011a,Burge:2013}. These
are used to train probabilistic models that score the likelihood that
a database sequence is generated by the same evolutionary processes as
the training sequences based upon both sequence and structural
information \cite{Sakakibara:1994,Eddy:1994,Nawrocki:2009}.  The
tRNAscan-SE software package uses CMs to accurately predict transfer
RNAs \cite{Lowe:1997,Chan:2009}.

@@

The conservation of these ``human'' lncRNAs among birds suggests they
may also be functional in birds. But what these functions are, is not
immediately obvious. For example, PART1 and PCA3 are both described as
prostate-specific lncRNAs that play a role in the human
androgen-receptor pathway
\cite{Bussemakers:1999,Lin:2000,Ferreira:2012}. Birds lack a prostate
but both males and females express the androgen receptor (AR or NR3C4)
in gonadal and non- gonadal tissue
\cite{Yoshimura:1993,Veney:2004,Fuxjager:2012,Leska:2012}. Thus, we
postulate that PART1 and PCA3 also play a role in the
androgen-receptor pathway in birds but whether the expression of these
lncRNAs are tissue specific is unknown at present.

@@

Vertebrate Y RNAs typically form a cluster comprising four
well-defined paralog groups Y1, Y3, Y4, and Y5. In line with
\cite{Mosig:07a} we find that the Y5 paralog family is absent from all
bird genomes, while it is still present in both alligator and turtle,
see Supplemental Figure~4. Within the avian lineage, we find a
conserved Y4-Y3-Y1 cluster. Apparently, broken-up clusters are in most
cases consistent with breaks (e.g. ends of contigs) in the available
sequence assemblies.  In several genomes we observe one or a few
additional Y RNA homologs unlinked to the canonical Y RNA
cluster. These sequences can be identified unambiguously as derived
members of one of the three ancestral paralog groups, they almost
always fit less well to the consensus (as measured by the CM bit score
of paralog group specific covariance models) than the paralog linked
to cluster, and there is no indication that any of these additional
copies is evolutionarily conserved over longer time scales. We
therefore suggest that most or all of these interspersed copies are in
fact pseudogenes.

@@

Seven families of ncRNAs were found in some avian genomes but not
others (Figure~\ref{fig:1}). These families range in conservation level
from being ubiquitous to cellular-life (RNase P and tRNA-sec), present
in most Bilateria (vault), present in the majority of eukaryotes
(RNase MRP, U4atac and U11) and present in all vertebrates
(telomerase) \cite{Hoeppner:2012}. Therefore, the genuine loss or even
diversification of these ncRNA families in the avian lineage is
unlikely. Rather, this lack of phylogenetic signal, combined with the
fragmented nature of the vast majority of these genomes described
above (i.e., of the 48 avian genomes, only the chicken and zebra finch
are chromosomally assembled \cite{birds:14,Warren:2010}), suggests the most likely
explanation is that these ncRNA families are indicative of missing
data. Indeed, of the seven missing ncRNA families, six where found in
the chicken genome and three were found in the zebra finch
genome. Furthermore, only one of these (RNase MRP) is found on a
macrochromosome, and all remaining missing ncRNAs are found on
microchromosomes (see Supplemental Table ?). A Fisher’s exact test
showed that there is significantly more missing ncRNAs on
microchromosomes than macrochromosomes ($P<10^{16}$ for both the
chicken and zebra finch). Thus, we suggest that many of
these ncRNAs families are missing because: (1) they are predominantly
found on microchromosomes [this study] and (2) the vast majority of
avian microchromosomes remain unassembled \cite{birds:14,Ellegren:2005}.

@@

Heatmaps showing the presence/absence and approximate genomic
copy-number of ``unusually, well conserved RNAs'' (particularly the
lncRNAs) on the left and families that have been identified as RNA
losses, divergence or missing data. In several cases functionally
related families have also been included, e.g. the RNA components of
the major and minor spliceosomes: U1, U2, U4, U5 and U6; and U11, U12,
U4atac, U5 and U6atac, respectively.\label{fig:1}

 \begin{figure}[ht]
   \centering
   \includegraphics[width=0.95\textwidth]{figures/figure1.pdf}
 \end{figure}

@@

This table contains the total number of annotated ncRNAs from
different RNA types in human, the median number for each of the 48
birds and chicken. The number of chicken ncRNA that show evidence for
expression is also indicated (the percentage is given in parentheses).
The threshold for determining expression was selected based upon a
false positive rate of less than 10\%.\label{table:1}

@@

\begin{tabular}{|r|r|r|r|l|}
\hline 
\multicolumn{5}{|l|}{{\bf ncRNA genes in human, chicken and all bird genomes}}\\
\hline 
                &                  &                   & Chicken ncRNAs & \\
Number in human & median(48 birds) & Number in chicken & confirmed with RNA-seq & RNA type\\
\hline
62&25.0&34&12 (35.3\%) &Long non-coding RNA\\ 
356&499.5&427&280 (65.6\%) &microRNA\\ 
281&120.0&106&90 (84.9\%) &C/D box snoRNA\\ 
336&85.5&68&48 (70.6\%) &H/ACA box snoRNA\\ 
34&13.0&12&12 (100.0\%) &Small cajal body RNA\\ 
1754&48.5&71&32 (45.1\%) &Major spliceosomal RNA\\ 
58&3.0&6&3 (50.0\%) &Minor spliceosomal RNA\\ 
525&82.0&122&88 (72.1\%) &Cis-regulatory element\\ 
316&6.5&9&3 (33.3\%) &7SK RNA\\ 
1&0.0&2&0 (0.0\%) &Telomerase RNA\\ 
9&0.0&2&1 (50.0\%) &Vault RNA\\ 
892&3.0&3&2 (66.7\%) &Y RNA\\ 
1084&173.5&300&278 (92.7\%) &Transfer RNA\\ 
80&9.5&4&2 (50.0\%) &Transfer RNA pseudogene\\ 
941&3.0&4&2 (50.0\%) &SRP RNA\\ 
607&7.0&22&10 (45.5\%) &Ribosomal RNA\\ 
4&1.0&2&2 (100.0\%) &RNase P/MRP RNA\\ 
\hline
7340&1080.0&1194&865 (72.4\%) &Total\\ 
\hline
\end{tabular}

>>>>

\begin{abstract}
Here we present the results of a large-scale bioinformatic annotation
of non-coding RNA loci in 48 avian genomes. Our approach uses
probabilistic models of hand-curated families from the Rfam database
to infer conserved RNA families within each avian genome. We
supplement these annotations with predictions from the tRNA annotation
tool, tRNAscan-SE and microRNAs from miRBase.  We show a number of
lncRNA-associated loci are conserved between birds and mammals,
including several intriguing cases where the reported mammalian lncRNA
function is not conserved in birds.  We also demonstrate extensive
conservation of classical ncRNAs (e.g., tRNAs) and more recently
discovered ncRNAs (e.g., snoRNAs and miRNAs) in birds. Furthermore, we
describe numerous ``losses'' of several RNA families, and attribute
these to genuine loss, divergence or missing data.  In particular, we
show that many of these losses are due to the challenges associated
with assembling Avian microchromosomes. These combined results illustrate the utility of
applying homology-based methods for annotating novel vertebrate
genomes.
\end{abstract}

@@

New sequencing technologies have dramatically expanded the rate at
which ncRNAs are discovered and their functions are determined
\cite{cech2014noncoding}. However, in order to determine the full
range of ncRNAs across multiple species we require multiple RNA
fractions (e.g. long and short), in multiple species, in multiple
developmental stages and tissues types. The costs of this approach are
still prohibitive in terms of researcher-time and
finances. Consequently, in this study we concentrate on bioinformatic
approaches, primarily homology based methods (i.e. covariance models
(CMs)). However, we do validate the majority of these predictions
using RNA-seq. The bioinformatic approaches we use remain state of the art for ncRNA bioinformatic
analyses \cite{Sakakibara:1994,Eddy:1994,Nawrocki:2009} and have well
established sensitivity and specificity rates \cite{Freyhult:2007}.
For example, the CM based approach for annotating ncRNAs in genomes requires
reliable alignments and consensus secondary structures of
representative sequences of RNA families, many of which can be found
at Rfam
\cite{Griffiths-Jones:2003,Griffiths-Jones:2005,Gardner:2009,Gardner:2011a,Burge:2013}. These
are used to train probabilistic models that score the likelihood that
a database sequence is generated by the same evolutionary processes as
the training sequences based upon both sequence and structural
information \cite{Sakakibara:1994,Eddy:1994,Nawrocki:2009}.  The
tRNAscan-SE software package uses CMs to accurately predict transfer
RNAs \cite{Lowe:1997,Chan:2009}.

@@

The conservation of these ``human'' lncRNAs among birds suggests they
may also be functional in birds. But what these functions may be is not
immediately obvious. For example, PART1 and PCA3 are both described as
prostate-specific lncRNAs that play a role in the human
androgen-receptor pathway
\cite{Bussemakers:1999,Lin:2000,Ferreira:2012}. Birds lack a prostate
but both males and females express the androgen receptor (AR or NR3C4)
in gonadal and non- gonadal tissue
\cite{Yoshimura:1993,Veney:2004,Fuxjager:2012,Leska:2012}. Thus, we
postulate that PART1 and PCA3 also play a role in the
androgen-receptor pathway in birds but whether the expression of these
lncRNAs are tissue specific is unknown at present.

@@

Vertebrate Y RNAs typically form a cluster comprising four
well-defined paralog groups Y1, Y3, Y4, and Y5. In line with
\cite{Mosig:07a} we find that the Y5 paralog family is absent from all
bird genomes, while it is still present in both alligator and turtle,
see Supplemental Figure~4. Within the avian lineage, we find a
conserved Y4-Y3-Y1 cluster. Apparently, broken-up clusters are in most
cases consistent with breaks (e.g. ends of contigs) in the available
sequence assemblies.  In several genomes we observe one or a few
additional Y RNA homologs unlinked to the canonical Y RNA
cluster. These sequences can be identified unambiguously as derived
members of one of the three ancestral paralog groups, they almost
always fit less well to the consensus (as measured by the CM bit score
of paralog group specific covariance models) than the paralog linked
to cluster, and there is no indication that any of these additional
copies is evolutionarily conserved over longer time scales. We
therefore suggest that most or all of these interspersed copies are in
fact pseudogenes (see below).

@@

Seven families of ncRNAs were found in some avian genomes but not
others (Figure~\ref{fig:1}). These families range in conservation level
from being ubiquitous to cellular-life (RNase P and tRNA-sec), present
in most Bilateria (vault), present in the majority of eukaryotes
(RNase MRP, U4atac and U11) and present in all vertebrates
(telomerase) \cite{Hoeppner:2012}. Therefore, the genuine loss or even
diversification of these ncRNA families in the avian lineage is
unlikely. Rather, this lack of phylogenetic signal, combined with the
fragmented nature of the vast majority of these genomes described
above (i.e., of the 48 avian genomes, only the chicken and zebra finch
are chromosomally assembled \cite{birds:14,Warren:2010}), suggests the most likely
explanation is that these ncRNA families are indicative of missing
data. Indeed, of the seven missing ncRNA families, six where found in
the chicken genome and three were found in the zebra finch
genome. Furthermore, only one of these (RNase MRP) is found on a
macrochromosome, and all remaining missing ncRNAs are found on
microchromosomes (see Supplemental Table~1). A Fisher’s exact test
showed that there is significantly more missing ncRNAs on
microchromosomes than macrochromosomes ($P<10^{16}$ for both the
chicken and zebra finch). Thus, we suggest that many of
these ncRNAs families are missing because: (1) they are predominantly
found on microchromosomes [this study] and (2) the vast majority of
avian microchromosomes remain unassembled \cite{birds:14,Ellegren:2005}.

@@

Heatmaps showing the presence/absence and approximate genomic
copy-number of ``unusually, well conserved RNAs'' (particularly the
lncRNAs) on the left and families that have been identified as RNA
losses, divergence or missing data. In several cases functionally
related families have also been included, e.g. the RNA components of
the major and minor spliceosomes: U1, U2, U4, U5 and U6; and U11, U12,
U4atac, U5 and U6atac, respectively.

 \begin{figure}[ht]
   \centering
   \includegraphics[width=0.95\textwidth]{figures/figure1.pdf}
   \caption{}
   \label{fig:1}
 \end{figure}

@@

This table contains the total number of annotated ncRNAs from
different RNA types in human, the median number for each of the 48
birds and chicken. The number of chicken ncRNA that show evidence for
expression is also indicated (the percentage is given in parentheses).
The threshold for determining expression was selected based upon a
false positive rate of less than 10\%.

@@

\begin{table}
\begin{tabular}{|r|r|r|r|l|}
\hline 
\multicolumn{5}{|l|}{{\bf ncRNA genes in human, chicken and all bird genomes}}\\
\hline 
                &                  &                   & Chicken ncRNAs & \\
Number in human & median(48 birds) & Number in chicken & confirmed with RNA-seq & RNA type\\
\hline
62&25.0&34&12 (35.3\%) &Long non-coding RNA\\ 
356&499.5&427&280 (65.6\%) &microRNA\\ 
281&120.0&106&90 (84.9\%) &C/D box snoRNA\\ 
336&85.5&68&48 (70.6\%) &H/ACA box snoRNA\\ 
34&13.0&12&12 (100.0\%) &Small cajal body RNA\\ 
1754&48.5&71&32 (45.1\%) &Major spliceosomal RNA\\ 
58&3.0&6&3 (50.0\%) &Minor spliceosomal RNA\\ 
525&82.0&122&88 (72.1\%) &Cis-regulatory element\\ 
316&6.5&9&3 (33.3\%) &7SK RNA\\ 
1&0.0&2&0 (0.0\%) &Telomerase RNA\\ 
9&0.0&2&1 (50.0\%) &Vault RNA\\ 
892&3.0&3&2 (66.7\%) &Y RNA\\ 
1084&173.5&300&278 (92.7\%) &Transfer RNA\\ 
80&9.5&4&2 (50.0\%) &Transfer RNA pseudogene\\ 
941&3.0&4&2 (50.0\%) &SRP RNA\\ 
607&7.0&22&10 (45.5\%) &Ribosomal RNA\\ 
4&1.0&2&2 (100.0\%) &RNase P/MRP RNA\\ 
\hline
7340&1080.0&1194&865 (72.4\%) &Total\\ 
\hline
\end{tabular}
\caption{}
\label{table:1}
\end{table}
